Defining Transcriptome using RNA Sequencing\\ of {\em Streptococcus mutans}

### 2012-01-21 package/manuscript.Rnw
Just an update.

<<echo=FALSE,result=hide>>=
glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
                            condA="ua159", condB="tw1" )
smutans.de2List( glucoseOnly, "glucoseOnly.csv" )
@
% Differential expression study
We sequenced 
\Sexpr{length(colnames(counts(smutansGenes)))}
RNA samples for studying differential gene expression to better understand
biological processes. Two independent factors of the biological processes
includes the effect of CcpA and the treatment of monosaccharides (glucose versus
galactose). The two factors results four pair-wise comparisons. We compared two
conditions of one factor with the other factor being fixed at one condition. For
example, we could compare UA159 wild-type and TW1 mutant-type strains under the
condition of glucose-grown cells to find differentially expressed associatively
with deactivation of \textit{ccpA} genes. We performed a similar comparison with
galactose-grown cells. While we replicated four biological experiments for the
condition combination of UA159 wilde-type grown in galactose, we had three for
the other three conditions. Because of limited resources of RNA deep sequencing,
we considered three replicates as a maximum number to but still adequately
detect differentially expressed genes.

We list differentially expressed genes in the comparison of UA159 wild-type and
TW1 mutant-type in Table \ref{tabGlucoseOnly}.

>>>>

<<echo=FALSE,result=hide>>=
#glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
#glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
#                            condA="ua159", condB="tw1" )
#save(glucoseOnly, file=file.path("glucoseOnly.RData"))
load("glucoseOnly.RData")
smutans.de2List( glucoseOnly, "glucoseOnly.csv" )
tableGlucoseOnly <- smutans.de2List( glucoseOnly, qval=1e-3 )

# Gene descriptions
geneDescription <- read.delim("feature-genome.out-geneonly.txt",head=FALSE,sep="\t")
geneDescription <- geneDescription[,c(4,7)]
colnames(geneDescription) <- c("Gene ID","Gene Description")
# Functional classes
functionalDescription <- read.csv("feature-genome.out-geneonly-oralgen-functional.txt",head=FALSE)
functionalDescription <- functionalDescription[,c(2,12)]
colnames(functionalDescription) <- c("Gene ID","Functional Class")
# Find potential differentially expressed genes, and add gene descriptions, and
# functional classes
x <- smutans.de2List( glucoseOnly, qval=1e-3 )
x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
tableGlucoseOnly <- x
@

% Differential expression study
We sequenced 
\Sexpr{length(colnames(counts(smutansGenes)))}
RNA samples for studying differential gene expression to better understand
biological processes. Two independent factors of the biological processes
includes the effect of CcpA and the treatment of monosaccharides (glucose versus
galactose). The two factors results four pair-wise comparisons. We compared two
conditions of one factor with the other factor being fixed at one condition. For
example, we could compare UA159 wild-type and TW1 mutant-type strains under the
condition of glucose-grown cells to find differentially expressed associatively
with deactivation of \textit{ccpA} genes. We performed a similar comparison with
galactose-grown cells. While we replicated four biological experiments for the
condition combination of UA159 wilde-type grown in galactose, we had three for
the other three conditions. Because of limited resources of RNA deep sequencing,
we considered three replicates as a maximum number to but still adequately
detect differentially expressed genes.

We list differentially expressed genes in the comparison of UA159 wild-type and
TW1 mutant-type in Table \ref{tabGlucoseOnly}.  
Of \Sexpr{nrow(counts(smutansGenes))} genes, 
\Sexpr{nrow(tableGlucoseOnly)} genes were found to be differentially expressed
at FDR of 0.1\%.

### 2012-01-21 package/supplementary.Rnw
Just an update.

<<echo=FALSE,results=hide>>=
# not run when smutans R package is installed
# Use this when smutans R package is being tested.
library(DESeq)
library(GenomicRanges)
library(plotrix)
library(xtable)
source("smutans/R/class_and_slots.R")
source("smutans/R/core.R")
source("smutans/R/methods.R")
load("smutans/data/smutansData.tx.RData")
load("smutans/data/smutansData.txGenes.RData")
load("smutans/data/smutansData.txPileup.RData")
load("smutans/data/smutans.feature.genes.RData")
load("smutans/data/smutans.genes.criteria.RData")
load("smutans/data/smutans.go.genes.RData")
load("smutans/data/smutans.cat.desc.RData")
load("smutans/data/smutansGenes.RData")
@

@@

<<echo=FALSE,results=hide>>=
geneDescription <- read.delim("feature-genome.out-geneonly.txt",head=FALSE,sep="\t")
geneDescription <- geneDescription[,c(4,7)]
colnames(geneDescription) <- c("Gene ID","Gene Description")
@

<<echo=FALSE,results=tex>>=
#glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
#glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
#                            condA="ua159", condB="tw1" )
#save(glucoseOnly, file=file.path("glucoseOnly.RData"))
load("glucoseOnly.RData")
x <- smutans.de2List( glucoseOnly, qval=1e-3 )
x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x.big <- xtable( x[order(x[,"p-value"]),],
                 display=c("s","s","d","d","f","e","s"),
                 digits=0,
                 align=c('l','l','r','r','r','r','p{3.5in}'),
                 label='tabGlucoseOnly',
                 caption='List of genes for the comparison UA159 and TW1 under glocuse-grown cells',
               )
print( x.big,
       tabular.environment='longtable',
       caption.placement="top",
       include.rownames=FALSE,
       floating=FALSE) 
@
\clearpage{}

<<echo=FALSE,results=tex>>=
#galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
#galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
#                              condA="ua159", condB="tw1" )
#save(galactoseOnly, file=file.path("galactoseOnly.RData"))
load("galactoseOnly.RData")
x <- smutans.de2List( galactoseOnly, qval=1e-3 )
x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x.big <- xtable( x[order(x[,"p-value"]),],
                 display=c("s","s","d","d","f","e","s"),
                 digits=0,
                 align=c('l','l','r','r','r','r','p{3.5in}'),
                 label='tabGalactoseOnly',
                 caption='List of genes for the comparison UA159 and TW1 under galactose-grown cells',
               )
print( x.big,
       tabular.environment='longtable',
       caption.placement="top",
       include.rownames=FALSE,
       floating=FALSE) 
@
\clearpage{}

<<echo=FALSE,results=tex>>=
#ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
#ua159only <- smutans.de2( ua159only, type="ua159", 
#                          condA="glucose", condB="galactose" )
#save(ua159only, file=file.path("ua159only.RData"))
load("ua159only.RData")
x <- smutans.de2List( ua159only, qval=1e-3 )
x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","Glucose","Galactose","fold", "p-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x.big <- xtable( x[order(x[,"p-value"]),],
                 display=c("s","s","d","d","f","e","s"),
                 digits=0,
                 align=c('l','l','r','r','r','r','p{3.5in}'),
                 label='tabUA159Only',
                 caption='List of genes for the comparison of UA159 wild-type grown with glucose and galactose ',
               )
print( x.big,
       tabular.environment='longtable',
       caption.placement="top",
       include.rownames=FALSE,
       floating=FALSE) 
@
\clearpage{}

<<echo=FALSE,results=tex>>=
#tw1only <- newSmutans( smutansGenes, title="TW1 Only" )
#tw1only <- smutans.de2( tw1only, type="tw1", 
#                          condA="glucose", condB="galactose" )
#save(tw1only, file=file.path("tw1only.RData"))
load("tw1only.RData")
x <- smutans.de2List( tw1only, qval=1e-3 )
x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","Glucose","Galactose","fold", "p-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x.big <- xtable( x[order(x[,"p-value"]),],
                 display=c("s","s","d","d","f","e","s"),
                 digits=0,
                 align=c('l','l','r','r','r','r','p{3.5in}'),
                 label='tabTW1Only',
                 caption='List of genes for the comparison of TW1 mutant-type grown with glucose and galactose ',
               )
print( x.big,
       tabular.environment='longtable',
       caption.placement="top",
       include.rownames=FALSE,
       floating=FALSE) 
@
\clearpage{}

>>>>

<<echo=FALSE,results=hide>>=
# not run when smutans R package is installed
# Use this when smutans R package is being tested.
library(DESeq)
library(GenomicRanges)
library(plotrix)
library(xtable)
library(RColorBrewer)
source("smutans/R/class_and_slots.R")
source("smutans/R/core.R")
source("smutans/R/methods.R")
load("smutans/data/smutansData.tx.RData")
load("smutans/data/smutansData.txGenes.RData")
load("smutans/data/smutansData.txPileup.RData")
load("smutans/data/smutans.feature.genes.RData")
load("smutans/data/smutans.genes.criteria.RData")
load("smutans/data/smutans.go.genes.RData")
load("smutans/data/smutans.cat.desc.RData")
load("smutans/data/smutansGenes.RData")
@

@@

<<echo=FALSE,results=hide>>=
geneDescription <- read.delim("feature-genome.out-geneonly.txt",head=FALSE,sep="\t")
geneDescription <- geneDescription[,c(4,7)]
colnames(geneDescription) <- c("Gene ID","Gene Description")

functionalDescription <- read.csv("feature-genome.out-geneonly-oralgen-functional.txt",head=FALSE)
functionalDescription <- functionalDescription[,c(2,12)]
colnames(functionalDescription) <- c("Gene ID","Functional Class")
@

<<echo=FALSE,results=tex>>=
#glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
#glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
#                            condA="ua159", condB="tw1" )
#save(glucoseOnly, file=file.path("glucoseOnly.RData"))
load("glucoseOnly.RData")
x <- smutans.de2List( glucoseOnly, qval=1e-3 )
x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x.big <- xtable( x[order(x[,"p-value"]),],
                 display=c("s","s","d","d","f","e","s"),
                 digits=c(0,0,0,0,1,0,0),
                 align=c('l','l','r','r','r','r','p{3.5in}'),
                 label='tabGlucoseOnly',
                 caption='List of genes for the comparison UA159 and TW1 under glocuse-grown cells',
               )
print( x.big,
       tabular.environment='longtable',
       caption.placement="top",
       include.rownames=FALSE,
       floating=FALSE) 
x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
@
\clearpage{}

We confirmed that the most upregulated genes in the TW1 mutant-type encoded the
components of the PDH enzyme complex (SMU.1421, SMU.1422, SMU.1423, and
SMU.1424).  
\Sexpr{length(grep("Energy metabolism",x[,"Functional Class"]))}
\Sexpr{length(grep("PTS",x[,"Functional Class"]))}
\Sexpr{length(grep("hypothetical",x[,"Gene Description"]))}

<<figPieGlucoseOnly,fig=TRUE,echo=FALSE,width=6,height=8>>=
par(mai=c(0,0,0,0),fig=c(0,1,0,1),mfrow=c(2,1))
#par(mfrow=c(2,1))
trim <- function (x) gsub("^\\s+|\\s+$", "", x)
y <- x[,"Functional Class"]
z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
z <- table(as.factor(sort(unlist(z))))

slices <- as.vector(z)
lbls <- names(z)
pct <- round(slices/sum(slices)*100)
c1 <- brewer.pal(n=9,name="Set1")
pie(z,labels=pct,main="",col=c1,clockwise=TRUE)
plot(0,0,axes=FALSE,frame.plot=FALSE,xlab="",ylab="",cex=0)
legend(-0.8,1,names(z),bty='n',cex=0.7,fill=c1)
@
\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{supplementary-figPieGlucoseOnly}
\caption{{\bf Distribution of functional classes of differntially expressed genes
in comparing UA159 wild-type and TW1 mutant-type grown in glucose.}
Labels for functional categories are colored with nine different colors. The
first label with red color, 
``Biosynthesis of cofactors, prosthetic groups, and carriers; Thiamine'', starts
at the top, and next labels read to the clockwise. The numbers circling the pie
chart is in per cent.
}
\label{figPieGlucoseOnly}
\end{figure}

\clearpage{}
<<echo=FALSE,results=tex>>=
#galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
#galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
#                              condA="ua159", condB="tw1" )
#save(galactoseOnly, file=file.path("galactoseOnly.RData"))
load("galactoseOnly.RData")
x <- smutans.de2List( galactoseOnly, qval=1e-3 )
x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x.big <- xtable( x[order(x[,"p-value"]),],
                 display=c("s","s","d","d","f","e","s"),
                 digits=c(0,0,0,0,1,0,0),
                 align=c('l','l','r','r','r','r','p{3.5in}'),
                 label='tabGalactoseOnly',
                 caption='List of genes for the comparison UA159 and TW1 under galactose-grown cells',
               )
print( x.big,
       tabular.environment='longtable',
       caption.placement="top",
       include.rownames=FALSE,
       floating=FALSE) 
x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
@
\clearpage{}

<<figPieGalactoseOnly,fig=TRUE,echo=FALSE,width=6,height=8>>=
par(mai=c(0,0,0,0),fig=c(0,1,0,1),mfrow=c(2,1))
#par(mfrow=c(2,1))
trim <- function (x) gsub("^\\s+|\\s+$", "", x)
y <- x[,"Functional Class"]
z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
z <- table(as.factor(sort(unlist(z))))

slices <- as.vector(z)
lbls <- names(z)
pct <- round(slices/sum(slices)*100)
c1 <- brewer.pal(n=9,name="Set1")
pie(z,labels=pct,main="",col=c1,clockwise=TRUE)
plot(0,0,axes=FALSE,frame.plot=FALSE,xlab="",ylab="",cex=0)
legend(-0.8,1,names(z),bty='n',cex=0.7,fill=c1)
@
\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{supplementary-figPieGalactoseOnly}
\caption{{\bf Distribution of functional classes of differntially expressed genes
in comparing UA159 wild-type and TW1 mutant-type grown in galactose.}
Labels for functional categories are colored with nine different colors. The
first label with red color, 
``Amino acid biosynthesis; Glutamate family'', starts
at the top, and next labels read to the clockwise. The numbers circling the pie
chart is in per cent.
}
\label{figPieGalactoseOnly}
\end{figure}

\clearpage{}
<<echo=FALSE,results=tex>>=
#ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
#ua159only <- smutans.de2( ua159only, type="ua159", 
#                          condA="glucose", condB="galactose" )
#save(ua159only, file=file.path("ua159only.RData"))
load("ua159only.RData")
x <- smutans.de2List( ua159only, qval=1e-10 )
x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","Glucose","Galactose","fold", "p-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x.big <- xtable( x[order(x[,"p-value"]),],
                 display=c("s","s","d","d","f","e","s"),
                 digits=c(0,0,0,0,1,0,0),
                 align=c('l','l','r','r','r','r','p{3.5in}'),
                 label='tabUA159Only',
                 caption='List of genes for the comparison of UA159 wild-type grown with glucose and galactose. Note that we used 1e-10 as the p-value cutoff.',
               )
print( x.big,
       tabular.environment='longtable',
       caption.placement="top",
       include.rownames=FALSE,
       floating=FALSE) 
x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
@
\clearpage{}

<<figPieUA159Only,fig=TRUE,echo=FALSE,width=6,height=8>>=
par(mai=c(0,0,0,0),fig=c(0,1,0,1),mfrow=c(2,1))
trim <- function (x) gsub("^\\s+|\\s+$", "", x)
y <- x[,"Functional Class"]
z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
z <- table(as.factor(sort(unlist(z))))

slices <- as.vector(z)
lbls <- names(z)
pct <- round(slices/sum(slices)*100)
c1 <- brewer.pal(n=9,name="Set1")
pie(z,labels=pct,main="",col=c1,clockwise=TRUE)
plot(0,0,axes=FALSE,frame.plot=FALSE,xlab="",ylab="",cex=0)

legend(-0.8,1,names(z),bty='n',cex=0.7,fill=c1)
#legend(-1.1,1.1,names(z[1:27]),bty='n',cex=0.7,fill=c1)
#legend(0,0.4,names(z[-c(1:27)]),bty='n',cex=0.7,fill=c1)

@
\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{supplementary-figPieUA159Only}
\caption{{\bf Distribution of functional classes of differntially expressed genes
in comparing UA159 wild-type grown in glucose and galactose.}
Labels for functional categories are colored with nine different colors. The
first label with red color, 
``Amino acid biosynthesis; Glutamate family'', starts
at the top, and next labels read to the clockwise. The numbers circling the pie
chart is in per cent.
}
\label{figPieUA159Only}
\end{figure}

\clearpage{}
<<echo=FALSE,results=tex>>=
#tw1only <- newSmutans( smutansGenes, title="TW1 Only" )
#tw1only <- smutans.de2( tw1only, type="tw1", 
#                          condA="glucose", condB="galactose" )
#save(tw1only, file=file.path("tw1only.RData"))
load("tw1only.RData")
x <- smutans.de2List( tw1only, qval=1e-3 )
x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","Glucose","Galactose","fold", "p-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x.big <- xtable( x[order(x[,"p-value"]),],
                 display=c("s","s","d","d","f","e","s"),
                 digits=c(0,0,0,0,1,0,0),
                 align=c('l','l','r','r','r','r','p{3.5in}'),
                 label='tabTW1Only',
                 caption='List of genes for the comparison of TW1 mutant-type grown with glucose and galactose ',
               )
print( x.big,
       tabular.environment='longtable',
       caption.placement="top",
       include.rownames=FALSE,
       floating=FALSE) 
x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
@
\clearpage{}

<<figPieTW1Only,fig=TRUE,echo=FALSE,width=6,height=8>>=
par(mai=c(0,0,0,0),fig=c(0,1,0,1),mfrow=c(2,1))
trim <- function (x) gsub("^\\s+|\\s+$", "", x)
y <- x[,"Functional Class"]
z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
z <- table(as.factor(sort(unlist(z))))

slices <- as.vector(z)
lbls <- names(z)
pct <- round(slices/sum(slices)*100)
c1 <- brewer.pal(n=9,name="Set1")
pie(z,labels=pct,main="",col=c1,clockwise=TRUE)
plot(0,0,axes=FALSE,frame.plot=FALSE,xlab="",ylab="",cex=0)
legend(-0.8,1,names(z),bty='n',cex=0.7,fill=c1)
@
\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{supplementary-figPieTW1Only}
\caption{{\bf Distribution of functional classes of differntially expressed genes
in comparing TW1 mutant-type grown in glucose and galactose.}
Labels for functional categories are colored with nine different colors. The
first label with red color, 
``Amino acid biosynthesis; Histidine family'', starts
at the top, and next labels read to the clockwise. The numbers circling the pie
chart is in per cent.
}
\label{figPieTW1Only}
\end{figure}
